---
title: "lubridate dates and times"
author: "Katherine Piatti"
date: "3/29/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# WORKING DATES AND TIMES

lubridate (in tidyverse) is the pkg for working with dates a times in r.


```{r}
library(tidyverse)
library(lubridate)

# get nyc flights data
library(nycflights13)
```




# today's date and time

```{r}
# get today's date
today()
```

```{r}
#get current time and date
now()
```



# convert strings to dates

```{r}
# convert strings into dates by specifying the order of the year, month, day

#year, month, day
ymd('2018-01-21')
```

```{r}
#month, day, year
mdy('jan 31st, 2019')
```

```{r}
# day, month, year
dmy('31-jan-17')
```



## convert numeric strings

```{r}
# above functions also take unquoted numbers
ymd(20171218)
```




# add time
```{r}

ymd_hms('2017-01-08 20:11:33')
```



## add timezone
```{r}
ymd(20171118, tz ='UTC')
```


# split across multiple columns

```{r}
flights %>% 
  select(year, month, day, hour, minute)
#> # A tibble: 336,776 x 5
#>    year month   day  hour minute
#>   <int> <int> <int> <dbl>  <dbl>
#> 1  2013     1     1     5     15
#> 2  2013     1     1     5     29
#> 3  2013     1     1     5     40
#> 4  2013     1     1     5     45
#> 5  2013     1     1     6      0
#> 6  2013     1     1     5     58
#> # … with 336,770 more rows
```

```{r}
# to create date/time vars from data split over multiple columns
flights %>% 
  select(year, month, day, hour, minute) %>% 
  mutate(departure = make_datetime(year, month, day, hour, minute))
```

```{r}
# (from "r for datascience", author is using below method b/c times in original dataset are represented in "slightly odd format"
make_datetime_100 <- function(year, month, day, time) {
  make_datetime(year, month, day, time %/% 100, time %% 100)
}

flights_dt <- flights %>% 
  filter(!is.na(dep_time), !is.na(arr_time)) %>% 
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) %>% 
  select(origin, dest, ends_with("delay"), ends_with("time"))

flights_dt
#> # A tibble: 328,063 x 9
#>   origin dest  dep_delay arr_delay dep_time            sched_dep_time     
#>   <chr>  <chr>     <dbl>     <dbl> <dttm>              <dttm>             
#> 1 EWR    IAH           2        11 2013-01-01 05:17:00 2013-01-01 05:15:00
#> 2 LGA    IAH           4        20 2013-01-01 05:33:00 2013-01-01 05:29:00
#> 3 JFK    MIA           2        33 2013-01-01 05:42:00 2013-01-01 05:40:00
#> 4 JFK    BQN          -1       -18 2013-01-01 05:44:00 2013-01-01 05:45:00
#> 5 LGA    ATL          -6       -25 2013-01-01 05:54:00 2013-01-01 06:00:00
#> 6 EWR    ORD          -4        12 2013-01-01 05:54:00 2013-01-01 05:58:00
#> # … with 328,057 more rows, and 3 more variables: arr_time <dttm>,
#> #   sched_arr_time <dttm>, air_time <dbl>
```
```{r}
# plot distibution of departure time across the year
flights_dt %>% 
  ggplot(aes(dep_time)) +
  geom_freqpoly(binwidth = 86400) #86400 seconds = 1 day
```


```{r}
# plot departure time across a single day
flights_dt %>% 
  filter(dep_time < ymd(20130102)) %>% 
  ggplot(aes(dep_time)) + 
  geom_freqpoly(binwidth = 600) # 600 s = 10 minutes
```
Note that when you use date-times in a numeric context (like in a histogram), 1 means 1 second, so a binwidth of 86400 means one day. For dates, 1 means 1 day.










======================================================================
R EXERCISES: lubridate exercises (part 1)
======================================================================

Exercise 1
Populate a variable called “start_date” with a date representation of string “23012017”

```{r}
(start_date <- dmy('23012017'))
```


Exercise 2
Use the lubridate function today to print the current date
```{r}
today()
```


Exercise 3
Extract the year part from the “start_date” variable created on exercise 1
```{r}
year(start_date)
```



Exercise 4
Extract the month part from the “start_date” variable created on exercise 1
```{r}
month(start_date)
```




Exercise 5
Extract the day part from the “start_date” variable created on exercise 1
```{r}
day(start_date)
```





Exercise 6
Set the month in variable “start_date” to February
```{r}
month(start_date) <- 2
start_date
```






Exercise 7
Add 6 days to variable “start_date”.
Did you notice what happened to the month value?
```{r}
start_date + days(6)
```






Exercise 8
Substract 3 months from variable “start_date”
```{r}
start_date - months(3)
```





Exercise 9 (Advanced)
Populate a field called concatenated_dates with a vector of dates containing the following values:
“31.12.2015”, “01.01.2016”, “15.02.2016”
```{r}
(concatenated_dates <- dmy(c("31.12.2015", "01.01.2016",  "15.02.2016")))
```





Exercise 10 (Advanced)
Calculate in a short and simple way the addition of 1 thru 10 days to “start_date” variable
```{r}
start_date + c(1:10) * days(1)
```



====================================================================
R EXERCISES: lubridate (part 2)
======================================================================

Exercise 1
Populate two variables called "start_date" and "end_date" with date representation of string “01/12/2015 15:40:32” and “01/10/2016 16:01:10”
```{r}
(start_date <- dmy_hms("01/12/2015 15:40:32"))
(end_date <- dmy_hms("01/10/2016 16:01:10"))
```





Exercise 2
Create an interval variable called my_interval based on variables “start_date” and “end_date”
```{r}
(my_interval <- interval(start_date, end_date))
```





Exercise 3
Show the class of the new object in order to make sure it’s an “interval”

Setting/Extracting date/time parts:
You can extract and set any part of a date/time variable using the following functions:
day(x)
mday(x)
wday(x, label = FALSE, abbr = TRUE)
qday(x)
yday(x)
hour(x)
minute(x)
second(x)

Feel free playing with the different options in the “wday” function. Of course, instead of extracting, you can set a specific value for each part of a date/time variable.
```{r}
class(my_interval)
```






Exercise 4
Extract the “week day” part from the “start_date” variable created on exercise 1
Display the name of the day in a long format.
```{r}
wday(start_date, label = T, abbr = F)
```






Exercise 5
Check if the “day of the year” in the “end_date” variable is greater than 230
```{r}
yday(end_date) > 230
```





Time Zones
Time zones represent the same “instant” across different geographic locations.
For example “2016-08-21 11:53:24” in UCT time zone is identical “2016-08-21 19:53:24” in CST TZ.
“lubridate” provides two functions that help dealing with time zones:
with_tz() : This function only changes the representation of the specific instant, but not its actual value force_tz() : This function changes the actual value of a specific instant.

Tip 1: use OlsonNames() to get a list of time zones defined in your machine
Tip 2: use grep("country_name",OlsonNames(),value=TRUE) to find a specific time zone

Exercise 6
Find the time zone representation for the city of “Buenos_Aires”
```{r}
grep('Buenos_Aires', OlsonNames(), value = T)
```




Exercise 7
Display the value of “end_date” variable according to “Buenos Aires” time zone
```{r}
with_tz(end_date, 'America/Buenos_Aires')
```


Exercise 8
Populate a variable called “NY_TZ” with the time zone value of “New_York” city
```{r}
(ny_tz <- grep('New_York', OlsonNames(), value = T))
```




Exercise 9
Set the time zone of “end_date” so it matches the time zone of New York. Populate a variable called “end_date_ny” with the result
```{r}
(end_date_ny <- force_tz(end_date, tzone = 'America/New_York'))
```





Exercise 10
Display the time difference between end_date_ny and end_date
```{r}
difftime(end_date_ny, end_date)
```


====================================================================
R EXERCISES lubridate (part 3)
=====================================================================

Exercise 1
Create a duration object having 260 seconds
```{r}
duration(num = 260, units = "seconds")
```
can also use dseconds(260)


Exercise 2
Create a duration object having 260 minutes and divide it by 60
```{r}
dminutes(260) / 60
```





Exercise 3
Create a duration object of 1 day length and substract from it 260 duration minutes
```{r}
oneday <- duration(num = 1, units = 'days')

oneday - dminutes(260)
```




Exercise 4
Create a duration object of 12 days, 5 hours and 10 minutes
```{r}
duration(day = 12, hour = 5, minute = 10)
```





Exercise 5
Create a sequence of periods, from zero to 10, by 2 weeks
```{r}
weeks(2) * c(0:10)
```




Exercise 6
round date ‘2016-10-16’ by ‘month’ unit
```{r}
round_date(ymd('2016-10-16'), unit = 'month')
```



Exercise 7
round date ‘2016-10-16’ down, by ‘month’ unit
```{r}
floor_date(ymd('2016-10-16'), unit = 'month')
```




Exercise 8
round date ‘2016-10-16’ up, by ‘month’ unit
```{r}
ceiling_date(ymd('2016-10-16'), unit = 'month')
```





Exercise 9 and 10
Calculating Thanksgiving Day in 2016:
Thanksgiving is celebrated on the fourth Thursday of November.
So we will execute the following steps:
1. Create a date/time object (date1 variable) representing Novevember 1st, 2016
2. Check which day of the week November 1st is.
3. Add the needed days to reach the upcoming Thursday (date2 variable)
4. Make sure your calculation is correct (e.g. you reached the upcoming Thursday)
5. Add 3 weeks
```{r}
date1 <- mdy('November 1st, 2016')
```

```{r}
wday(date1, label = T)
```

```{r}
date2 <- date1 + ddays(9)

wday(date2, label = T)
```








